Measurement Bias from Address Aliasing

### 2016-01-09 paper.tex
Many small fixes.

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
%Understanding program behavior and obtaining accurate measurements is important in performance analysis.
%However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
%Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
In this paper, we identify \emph{address aliasing} as an important underlying mechanism causing measurement bias on modern Intel microarchitectures.
By analyzing hardware performance counters, we show how bias can occur from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
%Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
%We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.

%This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
We demonstrate ways to deal with this type of bias, through runtime detection and correction of aliasing conditions.
This is especially important for heap allocations, as most allocators tend to produce worst case alignment by default; favoring page alignment.
%For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in already optimized code.
%Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
Our results can enable new architecture specific optimization strategies to account for this phenomenon.
\end{abstract}

@@

%Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture.
%We later reproduced similar results for the ``Core'', ``Nehalem'', and ``Ivy Bridge'' microarchitectures as well.
%Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.

@@

Best practices for controlling the execution context were applied, ensuring that we are not affected by any unwanted bias~\cite{Mytkowicz:2009:WrongData}.
Most importantly, this means keeping the memory address space under control.
For security reasons, addresses of the stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Pax:ASLR,Bhatkar:AddressObfuscation}. 
By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
All experiments are done with the machine under minimal load, disabling \emph{frequency scaling} to keep the CPU's clock speed fixed, and \emph{Hyper-threading} (HT) to reduce the possibility for resource contention between threads.\footnote{Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.}

>>>>

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
Understanding program behavior and obtaining accurate measurements is important in performance analysis.
However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.

This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
We demonstrate this by implementing runtime detection and correction of aliasing conditions.
For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in one of our examples.
Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
Our results can enable new architecture specific optimization strategies to account for this effect.
\end{abstract}

@@

Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture.
We later reproduced similar results for the ``Core'', ``Nehalem'', and ``Ivy Bridge'' microarchitectures as well.
Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.

@@

Best practices for controlling the execution context were applied, ensuring that we are not affected by any unwanted bias~\cite{Mytkowicz:2009:WrongData}.
Most importantly, this means keeping the memory address space under control.
For security reasons, addresses of the stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Pax:ASLR,Bhatkar:AddressObfuscation}. 
By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
All experiments are done on a dedicated machine under minimal load, with \emph{frequency scaling} disabled to keep the CPU's clock speed fixed.
Finally, we disable \emph{Hyper-threading} (HT), which reduces the possibility for resource contention between threads.
Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.
